---
title: 'Week 10: Temporal data'
date: "today"
output: pdf_document
execute:
  warning: no
  message: no
---

# Child mortality in Sri Lanka

In this lab you will be fitting a couple of different models to the data about child mortality in Sri Lanka, which was used in the lecture. Here's the data and the plot from the lecture:

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)

knitr::opts_chunk$set(echo=TRUE,
                      opts=list(width.cutoff=60),
                      tidy=TRUE,
                      fig.width=unit(10,"cm"),
                      fig.asp=0.5,
                      fig.align = "center",
                      cache = TRUE)

# theme_large_text <- theme(axis.text=element_text(size=15), axis.title=element_text(size=15))
```

```{r}
# lka <- read_csv(here("./lka.csv"))
lka <- read_csv("./lka.csv")
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(N = length(observed_years),
                  T = nyears,
                  mid_year = mean(years),
                  y = lka$logit_ratio,
                  se = lka$se,
                  years = years,
                  year_i = observed_years - years[1]+1)

# mod <- stan(data = stan_data,
#              file = here("code/models/lka_linear_me.stan"))

mod1 <- stan(data = stan_data,
             file = "./lka_linear_me.stan")

```

Extract the results:

```{r}
res1 <- mod1 %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```


Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res1, aes(year, .value)) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

## Question 1

Project the linear model above out to 2023 by adding a `generated quantities` block in Stan (do the projections based on the expected value $\mu$). Plot the resulting projections on a graph similar to that above. 

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(N=length(observed_years),
                  T = nyears,
                  mid_year = mean(years),
                  y = lka$logit_ratio,
                  se = lka$se,
                  years = years,
                  year_i = observed_years - years[1]+1,
                  P=9)
```

```{r}
mod2 <- stan(data = stan_data,
             file = "./lka_linear_me_2.stan")

```

Extract the outcome

```{r}
res2 <- mod2 |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year=years[t])

res_p <- mod2 |>
  gather_draws(mu_p[p]) |>
  median_qi() |>
  mutate(year=years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res1, aes(year, .value)) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+  geom_line(data = res_p, aes(year, .value)) + 
  geom_ribbon(data = res_p, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka",
       y = "log ratio", subtitle = "Linear fit shown in black")
```

# Random walks


## Question 2

Code up and estimate a first order random walk model to fit to the Sri Lankan data, taking into account measurement error, and project out to 2023. 

```{r}
mod_rw1 <- stan(data = stan_data,
             file = "./lka_linear_me_rw.stan")
mod_rw1
```

```{r}
res_rw1 <- mod_rw1 |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year=years[t])

res_prw1 <- mod_rw1 |>
  gather_draws(mu_p[p]) |>
  median_qi() |>
  mutate(year=years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes(color = source)) + 
  geom_line(aes(color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw() +
  geom_line(data = res_rw1, aes(year, .value)) + 
  geom_ribbon(data = res_rw1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_prw1, aes(year, .value), col="red") + 
  geom_ribbon(data = res_prw1, aes(y = .value, ymin = .lower, ymax = .upper), fill="red", alpha = 0.2) +
  theme_bw() +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Random walk fit shown in black, projections shown in red")
```

## Question 3

Now alter your model above to estimate and project a second-order random walk model (RW2). 

```{r}
mod_rw2 <- stan(data = stan_data,
             file = "./lka_linear_me_rw2.stan")
mod_rw2
```

```{r}
res_rw2 <- mod_rw2 |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year=years[t])

res_prw2 <- mod_rw2 |>
  gather_draws(mu_p[p]) |>
  median_qi() |>
  mutate(year=years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw() +
  geom_line(data = res_rw2, aes(year, .value)) + 
  geom_ribbon(data = res_rw2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_line(data = res_prw2, aes(year, .value),col="red") + 
  geom_ribbon(data = res_prw2, aes(y = .value, ymin = .lower, ymax = .upper),fill="red", alpha = 0.2) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Random walk fit shown in black, projections shown in red")
```

## Question 4

Run the first order and second order random walk models, including projections out to 2023. Compare these estimates with the linear fit by plotting everything on the same graph.

Use the outcome of the linear model from Q1, the first- and second-order random walk model from Q2 and Q3.

```{r}
ggplot(lka, aes(year, logit_ratio))  +
  geom_point() + 
  geom_line(lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se), alpha = 0.1) +
  geom_line(data = res1, aes(year, .value, col="Linear")) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper, fill="Linear"), alpha = 0.2) +
  geom_line(data = res_p, aes(year, .value, col="Linear")) + 
  geom_ribbon(data = res_p, aes(y = .value, ymin = .lower, ymax = .upper, fill="Linear"), alpha = 0.2) +
  theme_bw() +
  geom_line(data = res_rw1, aes(year, .value, col = "RW1")) +
  geom_ribbon(data = res_rw1, aes(y = .value, ymin = .lower, ymax = .upper, fill = "RW1"), alpha = 0.2) +
  geom_line(data = res_prw1, aes(year, .value,col="RW1")) +
  geom_ribbon(data = res_prw1, aes(y = .value, ymin = .lower, ymax = .upper,fill="RW1"), alpha = 0.2) +
   geom_line(data = res_rw2, aes(year, .value, col = "RW2")) + 
  geom_ribbon(data = res_rw2, aes(y = .value, ymin = .lower, ymax = .upper, fill = "RW2"), alpha = 0.2) +
 geom_line(data = res_prw2, aes(year, .value,col="RW2")) + 
 geom_ribbon(data = res_prw2, aes(y = .value, ymin = .lower, ymax = .upper,fill="RW2"), alpha = 0.2) +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Comparison of model fits with and without VR data (RW = random walk)", col = "Median fits",  fill = "Standard Error") +
  scale_color_manual(name="Model Fits", values=c("Linear"="black", "RW1"="red", "RW2"="blue"))
```


## Question 5

Rerun the RW2 model excluding the VR data. Briefly comment on the differences between the two data situations.

Need to filter out the VR data.

```{r}
# Filter out the VR data
lka_nVR <- lka %>% filter(source != "VR")
obs_years <- lka_nVR$year
years <- min(obs_years):max(obs_years)
nyears <- length(years)
```

```{r}
max(years)
```

After filtering out the VR data, the most recent year becomes 2005, therefore, we need to add in total 18 years into the projection.

```{r}
stan_data_nvr <- list(N=length(obs_years),
                      T = nyears,
                      mid_year = mean(years),
                      y = lka_nVR$logit_ratio,
                      se = lka_nVR$se,
                      years = years,
                      year_i = obs_years - years[1]+1,
                      P=18)
```


```{r}
mod_rw_nvr <- stan(data = stan_data_nvr,
                   file = "./lka_linear_me_rw2.stan",
                   cores=parallel::detectCores())
mod_rw_nvr
```

```{r}
res_rw_nvr <- mod_rw_nvr |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year=years[t])

res_prw_nvr <- mod_rw_nvr |>
  gather_draws(mu_p[p]) |>
  median_qi() |>
  mutate(year=years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio))  +
  geom_point() + 
  geom_line(lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se), alpha = 0.1) + 
  geom_line(data = res_rw_nvr, aes(year, .value, col = "No VR")) + 
  geom_ribbon(data = res_rw_nvr, aes(y = .value, ymin = .lower, ymax = .upper, fill = "No VR"), alpha = 0.2) +
 geom_line(data = res_prw_nvr, aes(year, .value,col="No VR projections")) + 
 geom_ribbon(data = res_prw_nvr, aes(y = .value, ymin = .lower, ymax = .upper,fill="No VR projections"), alpha = 0.2) +
  theme_bw() +
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio", subtitle = "Second-order random walk fits with and without VR data", col = "Source and Model Fits", fill = "Source and Model Fits")
```


## Question 6

Briefly comment on which model you think is most appropriate, or an alternative model that would be more appropriate in this context.

The linear model is the best here because it captures the upward trend of the temporal change of child mortality the best.

